Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-17 manuscript/paper.tex
formalism, overview para.

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages,
in the cubic structure, to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations (sec.~\ref{sec:rotation})
on
the structure (sec.~\ref{sec:structure}) 
and crystal field (sec.~\ref{sec:deformCF}).
The results are in sec.~\ref{sec:results},
where we seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (sec.~\ref{sec:t2g})
and $e_g$ (sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (sec.~\ref{sec:FL}).
\az{appendix?}

@@

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of the crystal field Hamiltonian of this system 
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed to the same~\cite{paxton-notes}.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The octahedral field at the centre of the octahedron or cube  
is given by~\cite{pavariniChap}, [include monopole?? bcs later we incl its changes...]
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B at the centre 
turns out to be diagonal
and splits the energy levels into two manifolds,
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

\section{Janh-Teller Effect}
\label{sec:JT}

>>>>

The paper is organised as follows.
Section~\ref{sec:formalistm}
presents the formalism and 
sec.~\ref{sec:JT} reviews the Jahn-Teller effect to
introduce the notation and concepts involved
by considering tetragonal distortion of an octahedron.
%\az{[.... multiple subsections?]}
The model and calculations are presented
in sec.~\ref{sec:calc},
where
we first discuss the relative contributions of the three closest ion cages,
in the cubic structure, to
the crystal field at the central transition metal site 
and crystal field splitting between its $e_g$ and $t_{2g}$ manifolds
(sec.~\ref{sec:relativeVoctABO}),
and then 
describe the deformation effects of 
rigid octahedral rotations (sec.~\ref{sec:rotation})
on
the structure (sec.~\ref{sec:structure}) 
and crystal field (sec.~\ref{sec:deformCF}).
The results are in sec.~\ref{sec:results},
where we %seperately
present the splitting in the single particle spectrum
belonging to the $t_{2g}$ (sec.~\ref{sec:t2g})
and $e_g$ (sec.~\ref{sec:eg-gen}) manifolds,
and finally describe the effect of the full lattice (sec.~\ref{sec:FL}).
\az{appendix?}

@@

Let us introduce the basic formalism we use throughout this manuscript.
The spectrum of an atom or ion
changes in the presence of nearby charged species.
This is described by the crystal field Hamiltonian of the atom
that contains the coupling between its atomic orbitals and/or their energy shifts
induced by the electrostatic potential of those charges.
Since the angular part of atomic orbitals are described by 
the spherical harmonics,
an expansion of the potential at the atom's site 
in the same bases simplify the analysis.
This is because
the effect of various multipole components of the potential
is governed by certain selection rules 
(enforced by the symmetries of the spherical harmonics
through Gaunt coefficients, see below).

If we have an arrangement of ions with charge $q$ around a given site,
we can express their (crystal field) potential in terms of spherical harmonics~(see \ref{append:multipoles}), 
\begin{align}
V(\vec r) &= \frac{q}{a}\sum_{lm}V_{l,m}~\left(\frac{r}{a}\right)^l \ylm{l,m},
\end{align}
and calculate the crystal field hamiltonian of the central ion/atom.
Assuming we only have d-orbitals ($l=2$) on the central atom,
the matrix elements of the crystal field Hamiltonian of this system 
are given by,
\begin{align}
H_{m',m''} &= -\frac{q}{a}\sum_{l,m} V_{l,m} C_{l}^{m'm''m} D_{l} ,
\end{align}
where 
$C_{l}^{m'm''m}=\braket{Y_{2,m'}| Y_{l,m}| Y_{2,m''}}$ are Gaunt coefficients for the given (real or complex) spherical harmonics,
and $D_{l}=\braket{(r/a)^l}$ is the expectation value for the d-orbitals, assumed to be the same for all d-orbitals, i.e., the radial part of all d-orbitals is assumed the same~\cite{paxton-notes}.

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 relevant for the d-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$)
 and called octahedral field. 
It is given by~\cite{pavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

\section{Jahn-Teller Effect}
\label{sec:JT}
